---
title: "Data wrangling"
author: "Laurel"
date: "October 22, 2018"
output: html_document
---



```{r}
library(tidyverse) #install.packages("tidyverse")
```

#Read data

This is the gapminder dataset.

```{r}
## read gapminder csv. Note the readr:: prefix identifies which package it's in
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')

#inspect dataset
gapminder #looks at all data
head(gapminder) #gives you the first 6 rows by default
tail(gapminder) #last observations, default is 6

head(gapminder, 3)
tail(gapminder, 10)

#summary statistics
str(gapminder) #structure of data: variables, char/num
summary(gapminder)

#install.packages("skimr")
skimr::skim(gapminder)

names(gapminder) #column names
dim(gapminder) #dimensions
ncol(gapminder) #number of columns
nrow(gapminder) #number of rows

#reverse-engineer dim
c(nrow(gapminder), ncol(gapminder)) #can commbine data steps

```


##Dplyr!

```{r}
#filter by rows
filter(gapminder, lifeExp < 29)
filter(gapminder, country == "Mexico")
filter(gapminder, country %in% c("Mexico", "Peru", "United States"))
filter(gapminder, country == "Mexico", year == 2002)

Sweden <- filter(gapminder, country == "Sweden")
mean(Sweden$lifeExp) #mean life expectancy for Sweden


```

```{r}
#select by columns
select(gapminder, year, lifeExp)
select(gapminder, -continent, -lifeExp) # - will exclude columns

#use select and filter together
gap_cambodia <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)

```

#Pipe operator

```{r}

gapminder %>% head(3)
# is equivalent to: 
head(gapminder, 3)

#let's improve this with pipes:
gap_cambodia <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)

gap_cambodia <- gapminder %>% 
  filter(country == "Cambodia") %>% 
  select(-continent, -lifeExp)

```

```{r}
#mutate adds new variables
gapminder %>% 
  mutate(index = 1:nrow(gapminder))
  

gapminder %>% 
  mutate(planet = "Earth")

gapminder %>% 
  mutate(gdp = pop * gdpPercap)

##example:
# Find the maximum GDP of Egypt and Vietnam. Create a new column. (I want just one number that is the max of those two countries)

gapminder %>% 
  filter(country %in% c("Egypt", "Vietnam")) %>%
  mutate(gdp = pop * gdpPercap) %>% 
  mutate(maxGDP = max(gdp))

#or

mutate(gdp = pop * gdpPercap,
       maxGDP = max(gdp))

```

#'group_by()' groups!

```{r}

## with mutate()
gap_grouped <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap,
         max_gdp = max(gdp)) %>% 
  ungroup()
#View(gap_grouped)

## with summarize() or summarise()

gap_summarized <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap) %>% 
  summarise(max_gdp = max(gdp)) %>% 
  ungroup()

#View(gap_summarized)

```

## 'arrange()' arranges columns

```{r}
gap_summarized <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap) %>% 
  summarise(max_gdp = max(gdp)) %>% 
  ungroup() %>% 
  arrange(desc(max_gdp))


#View(gap_summarized)
```

#Joining data

```{r}
## read in the data. (same URL as yesterday, with co2.csv instead of gapminder.csv)
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")

#explore
co2 %>% head()
co2 %>%  dim()

#create a new gapminder variable with only 2007 data
gap_2007 <- gapminder %>% 
  filter(year == 2007)

gap_2007 %>% dim() #142 observations

## left-join gap_2007 to co2
lj <- left_join(gap_2007, co2, by = "country") # have to put quotes around variable and have to have a space between = and variable error: object country not found if not using quotes

#another way to write this is:
lj <- gap_2007 %>% 
  left_join(co2, by = "country")

## right-join gap_2007 to co2
rj <- right_join(gap_2007, co2, by = "country")
# the same as but different column ordering
lj2 <- left_join(co2, gap_2007, by = "country")


```



```{r}
## read in gapminder data in wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')


```

## tidyr::gather()

- puttting a - in front of a variable name in the gather statement keeps the variable as is, not gatthered
- key is the column names, new name that you provide (eg. obstype_year)
- value is the values that were under the columns


```{r}
gap_long <- gap_wide %>% 
  gather(key = obstype_year,
         value = obs_values,
         -continent, -country)

```

## tidyr::separate()

- convert = TRUE converts all number columns to numeric


```{r}
gap_long <- gap_wide %>% 
  gather(key = obstype_year,
         value = obs_values,
         -continent, -country)  %>% 
  separate(col = obstype_year,
           into = c("obs_type", "year"),
           sep = "_", 
           convert = TRUE)

#str(gap_long)

```

## plot long data

```{r}
canada_df <- gap_long %>%
  filter(country == "Canada",
         obs_type == "lifeExp")

ggplot(data = canada_df, aes(x = year, y = obs_values)) +
  geom_line()
  
```

Plot the life expectancy of all countries in the Americas

```{r}
life_df <- gap_long %>% 
  filter(obs_type == "lifeExp",
         continent == "Americas")

ggplot(data = life_df, aes(x = year, y = obs_values, color = country)) +
  geom_line()
  
  
```

Exercise 7.5

Calculate and plot the mean life expectancy for each continent over time from 1982-2007.

```{r}
meanlife <- gap_long %>% 
 filter(obs_type == "lifeExp", 
year %in% c(1982:2007)) %>%
 group_by(continent, year) %>%  
  summarize(mean_life = mean(obs_values)) %>% 
  ungroup()
  
#plot using ggplot
ggplot(data= meanlife, aes(x = year, y = mean_life, color = continent)) +
  geom_line() +
  labs(x = "Year",
       y = "Age (years)",
       title = "Mean life expectancy") +
  theme_bw()


```

## spread()

- takes unique values in obs_type and makes a new column for each using obs_values

```{r}
gap_normal <- gap_long %>% 
  spread(key = obs_type,
         value = obs_values)


```

7.6 Exercise 

Convert back to wide data set

```{r}
gap_wide_new <- gap_long %>% 
  unite(col = names, obs_type, year, sep = "_") %>% 
  spread(key = names,
         value = obs_values)
```

